---
title: "cm012 Exercises: Factors"
output:
  html_document:
    keep_md: true
---

`forcats` package comes loaded with `tiyverse`:

```{r}
suppressPackageStartupMessages(library(tidyverse)) # tidyverse also loads forcats package
library(gapminder)
```

# Factors

## Resources

- Exercises are based on http://stat545.com/block029_factors.html and http://r4ds.had.co.nz/factors.html. Some content was taken from the former.

## Intro to Factors

What is a factor? A "truly categorical" variable. You can think of it as a vector that:

- has character entries on the surface
- are integers underneath
- has __levels__

Examples of Base R's obsession with coercing to factors:

```{r}
data.frame(x=c("A", "B")) %>% 
  str() # column x is a factor
```


```{r}
lotr1 <- "https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv" %>% 
  read.csv()
lotr2 <- "https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv" %>% 
  read.csv()
lotr1
str(lotr1)
str(lotr2)
```


Examples of problems encountered with factors. (ideas came from [R Bloggers](https://www.r-bloggers.com/factors-are-not-first-class-citizens-in-r/))

```{r}
head(iris)
iris %>% 
  mutate(Species = ifelse(Species == "versicolor", "vers", Species)) %>% 
  str() # R says no such level that is vers therefore it coerces it to a character vector
```

```{r}
c(iris$Species, "setosa") # again converts it into a character vector

as.character(iris$Species) # do what we want it to do!
```


- Base R~~ way of interacting with factors:
    - `factor()`, or `forcats::parse_factor()`.
    - `levels()`
    - `nlevels()`
    - `forcats::fct_count()`

Here is a sample of 10 letters drawn from the possibilities "a", "b", and "c":

```{r}
set.seed(10)
(draw <- sample(letters[1:3], size = 10, replace = TRUE))
```

Convert `draw` to a factor. What are the levels? How many are there? How many of each category was drawn?

```{r}
draw <- factor(draw)
draw

levels(draw)
nlevels(draw) %>% is.factor() # levels are not factors

# nice factor levels and count tibble
fct_count(draw)
```


## Concatenating Factors

We saw that `c()` doesn't work for concatenating. Modify the following code to use `fct_c()` from the `forcats` package:

```{r}
c(lotr1$Film, lotr2$Film)
```

Try binding by row `lotr1` and `lotr2`:

- with `rbind()`
- with `bind_rows()`

Which one is more lenient? Which would you prefer?

```{r}
rbind(lotr1, lotr2) %>% str() # film = factor w/2 levels (the film names)
bind_rows(lotr1, lotr2) %>%  str() # abandaon factor thing and goes to character for film
# if lotr1 and lotr2 had the same rows, then bind_rows would work as factor
as.character(5)
```


## Unused Levels

Levels don't always have to be present ("observed") in the factor. Example of what this means:

```{r}
gap_gs <- gapminder %>% 
  filter(country %in% c("Germany", "Sweden")) # filters to Germany and Sweden rows

levels(gap_gs$country) # keeps all 142 countries
nlevels(gap_gs$country)

as.character(gap_gs$country) # only results in Germany and Sweden
```


Sometimes keeping the levels is good. Other times, not.

__Example of when it's good__: 

Here's the gapminder data down to rows where population is less than a quarter of a million, i.e. 250,000:

```{r}
gap_small <- gapminder %>% 
  filter(pop < 250000)
```

Exercise: Make a bar chart of the number of times a continent has a country with population < 250,000 in the `gapminder` data set. Try with and without `scale_x_discrete(drop=FALSE)`.

```{r}
gap_small %>% 
  ggplot(aes(x = continent)) +
  geom_bar() + 
  scale_x_discrete(drop = FALSE) # don't drop Americas and Oceania even though they never had a pop < 250,000
```

__Example of when it's bad__: If you ever use the `levels()` function.

How to fix by dropping levels:

- Base R: `droplevels()` operates on either an entire data frame or a factor.
- `forcats::fct_drop()` only operates on a factor.
    
Exercise: get rid of the unused factor levels for country and continent in different ways:

- `droplevels()`
- `fct_drop()` inside `mutate()`
- Re-defining the variable as a factor

```{r}
gap_small %>% 
  droplevels() %>% # if no observation in the column, then it drops the corresponding level
  str() # continent only has levels Africa, Asia, and Europe and the countries of those guys

gap_small %>% 
  mutate(continent = fct_drop(continent)) %>% 
  str() # only have the levels of Africa, Asia and Europe (other continents, Americas and Oceania, are dropped b/c no observations in the column corresponding to pop < 250,000)
# still has 142 cuntries
```


## Ordering

Ordering of levels is alphabetical, by default. Usually not useful!

```{r}
cont <- gapminder$continent
levels(cont)
```

Plotting happens in the order of the factor levels:

```{r, fig.width=2, fig.height=1}
qplot(cont) # alphabetical order by continent by default
```

Much more effective to always consider a meaningful order when plotting a categorical variable. We'll look at three ways to re-order a factor. 

### Ordering with the factor itself

Reorder by frequency:

- Rearrange by frequency: `fct_infreq()`.
- Reverse: `fct_rev()` 

```{r, fig.width=2, fig.height=1}
# decreasing order of frequency
cont %>% 
  fct_infreq() %>% 
  qplot() 
# increasing order of frequency
cont %>% 
  fct_infreq() %>% 
  fct_rev() %>% 
  qplot()
```

Could also arrange by the order they appear in the factor with `fct_inorder()`.

```{r}
draw %>% 
  fct_inorder() # orders levels in order they appear in the factor
```


### Ordering by Another Variable

Here are the 2007 life expectancies of Asian countries:

```{r}
# further to the right means higher life expectancy
gap_asia_2007 <- gapminder %>% 
  filter(year == 2007, continent == "Asia")
ggplot(gap_asia_2007, aes(lifeExp, country)) + geom_point()
```

Let's use `fct_reorder()` to reorder the countries of `gap_asia_2007` by life Expectancy, and produce the same plot:

```{r, fig.width=4, fig.height=2}
gap_asia_2007 %>% 
  mutate(country = fct_reorder(country, lifeExp)) %>% # reorder countries by life Expectancy (more readable)
  ggplot(aes(lifeExp, country)) + geom_point() + 
  geom_point()
```

What about when life Expectancy is not unique? Example: life expectancy of each continent:

```{r, fig.width=2, fig.height=1}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_violin()


ggplot(gapminder, aes(continent, lifeExp)) +
  geom_boxplot()
```

`fct_reorder(f, x)` still works, but does some internal wrangling: a summary statistic (default: median) is computed on `x` for each category in the factor `f`.

Exercise: Try making the above box plot and violin plots, ordered by median lifeExp. Try other functions to order by by modifying the `.fun` argument.

```{r, fig.width=2, fig.height=1}
# ** fct_reorder() by default orders by the median lifeExp **
gapminder %>% 
mutate(contient = fct_reorder(continent, lifeExp)) %>% # reorder continents by median life Expectancy (more readable)
ggplot(aes(continent, lifeExp)) +
  geom_violin()


gapminder %>% 
mutate(continent = fct_reorder(continent, lifeExp)) %>% # reorder continents by median life Expectancy (more readable)
  ggplot(aes(continent, lifeExp)) +
  geom_boxplot()
```


What if we have two variables plus a non-positional categorical variable? Example: Life expectancy for some select countries. Want legend "ordered by life expectancy" -- but what does that mean?

```{r}
select_countries <- c("Sweden", "Denmark", "Norway", "Germany", "Austria")
gap_select <- gapminder %>%  # df of just the 5 countries
  filter(country %in% select_countries) %>% 
  droplevels() # drop unused levels
ggplot(gap_select, aes(year, lifeExp)) +
  geom_line(aes(group=country, colour=country)) # plot of life expectancy over time for thos 5 countries
```

Use `fct_reorder2(f, x, y)` to reorder factor `f`:

- `.fun` is a function of `x` and `y`. Should return a single value, and is applied to each category.
- Default is `.fun = last2`, which looks at x-y plot for each category; uses the y-value furthest to the right.

Exercise: Reorder the above line graph so that the legend is in order of last life expectancy. Useful for black-and-white printing!

```{r, fig.width=4, fig.height=2}
 # fct_reorder2(f, x, y) looks at max x value and takes corresponding y-value. Reorders the countries according to the life expectancy at the very end.
gap_select %>% 
  mutate(country = fct_reorder2(country, year, lifeExp)) %>% 
ggplot(aes(year, lifeExp)) +
  geom_line(aes(colour=country)) # plot of life expectancy over time for thos 5 countries

```

### Ordering "because I said so"

Remember the plot of Asian life expectancies in 2007? What if you're preparing a report for the Syrian government? You'd want to put Syria first (for reasons external to the data).

Here's how to use `fct_relevel()` to do that. Exercise: modify the code so that:

- in addition, Sweden goes second.
- instead of first, Syria goes after the third level. Hint: use `after=`.

```{r}
gap_asia_2007$country %>% 
  fct_relevel("Syria") %>% # releveling gap_asia_2007$country so that Syria goes first
  levels() %>% 
  head()

gap_asia_2007$country %>% 
  fct_relevel("Syria", "Sweden", after = 2) %>% # releveling gap_asia_2007$country so that Syria goes third (after 2) and Sweden goes 4th
  levels() %>% 
  head()

```

## Re-coding a Factor

Want "United States" to read "USA" instead? Just use `fct_recode()`. (Sadly, no metaprogramming happens here).

Exercise: modify the following code to also change "Canada" to read "Can". Hint: use a comma.

```{r}
gap_big_north <- gapminder %>% 
  filter(country %in% c("Canada", "United States", "Mexico")) %>% 
  droplevels()
gap_big_north$country %>% 
  fct_recode("USA" = "United States") %>%  # renames United States to USA
  levels()
```

## Condensing a Factor

We can specify levels to combine. Let's look at the world in 2007:

```{r}
gap_2007 <- gapminder %>% 
  filter(year == 2007)
```

We can arbitrarily combine levels using `fct_collapse()`. For example, combine Europe and Asia into one factor called "combo":

```{r, fig.width=4, fig.height=2}
cont %>% 
  fct_collapse("combo" = c("Europe", "Asia")) %>% # combo factor instead of Europe and Asia
  qplot()
```

More practically, we can lump the least frequent levels together as "Other". Modify the above code to use `fct_lump()` instead of `fct_collapse()` so that:

- The bar chart shows the two most frequently observed continents, 
- The bar chart shows the two least frequently observed continents (Hint: use negative `n`).
- You let `fct_lump()` decide on the number of non-other continents. How is this chosen?
- Note: you can manually specify non-other levels using `fct_other()`.

```{r, fig.width=2, fig.height=1}
cont %>% 
  fct_lump() %>% # by default, combines all the continents so that total combined height of the other bar is the smallest bar
  qplot()

# keep top two frequent levels and lump everything else together in 1
cont %>% 
  fct_lump(n=2) %>% # by default, combines all the continents so that total combined height of the bar is less than least frequent other factor
  qplot()
```

We can use the `w` argument to lump by another variable.

Exercise: Modify the following violin plot of life expectancies of African countries, so that:

1. There are 4 "violins" corresponding to countries with the highest lifeExp.
2. There are 4 "violins" corresponding to countries with the highest gdpPercap

```{r, fig.width=2, fig.height=1}
gap_africa <- gapminder %>% 
    filter(continent == "Africa")
gap_africa %>% 
    mutate(country = fct_lump(country)) %>% 
    ggplot(aes(country, lifeExp)) +
    geom_violin()
```


## Exercises

Use the `gss_cat` data to answer the following questions (from http://r4ds.had.co.nz/factors.html).

1. (15.3.1 Ex. 1) Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

```{r}

```

2. (15.3.2 Ex. 2) What is the most common relig in this survey? What’s the most common partyid?

```{r}

```

3. (15.5.1 Ex. 1) How have the proportions of people identifying as Democrat, Republican, and Independent changed over time? Modify the following plot to a friendlier legend order.

```{r, fig.width=4, fig.height=2}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>% 
  count(year, partyid) %>% 
  ggplot(aes(year, n)) +
  geom_line(aes(group=partyid, colour=partyid))
```

# Dates and Times with Lubridate

Goal here: some exposure to `lubridate`; know it exists.

1. Use different combinations of y, m, d to make a date time object.

```{r}
lubridate::ymd(170511)
lubridate::ymd("2017-May-11")
```

2. Get `year`, `month`, `yday`, `wday`, `day`.

```{r}

```

3. Add durations (exact time spans) with `ddays`, `dweeks`, ... and periods (human-interpretable time spans) with `days`, `weeks`, and especially `months`.

```{r}

```

